Using functor categories to generate intermediate code with Agda

### 2025-04-19 draft.tex
more definitions added

@@

    \section{Category theory background}
        Category 
        \subsection{Category}
        \begin{definition*}[Category]
            A category $\mathcal{C}$ is specified by the following:
            \begin{itemize}
                \item 
                    a collection of objects $\textbf{obj}(\mathcal{C})$, whose elements are called $\mathcal{C}$-objects;
                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a collection of morphisms $\mathcal{C}{(X,Y)}$, whose elements are called $\mathcal{C}$-morphisms from $X$ to $Y$;
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, an element $\textbf{id}_X \in \mathcal{C}{(X,X)}$ called the identity morphism on $X$;
                \item 
                    for each $X, Y, Z \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} \times \mathcal{C}{(Y,Z)} &\to \mathcal{C}{(X,Z)} \\
                        (f,g) &\mapsto g \circ f
                    \end{aligned}\]
                    called the composition of morphisms;
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    \textbf{(Unit Law)}
                    for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have:
                    \begin{equation} \label{law: unit}
                        \textbf{id}_Y \circ f = f = f \circ \textbf{id}_X
                    \end{equation}
                \item
                    \textbf{(Associativity Law)}
                    for all $X, Y, Z, W \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, $h \in \mathcal{C}{(Z,W)}$, we have:
                    \begin{equation} \label{law: associativity}
                        h \circ (g \circ f) = (h \circ g) \circ f
                    \end{equation}
            \end{itemize}

            [Considering adding different notations for morphisms, composition, identity morphisms, etc.]
            [Considering adding a diagram for the definition of category]
            [Considering adding examples of categories]

@@

        \subsection{Functor}
        \begin{definition*}[Functor]
            A functor $F$ from a category $\mathcal{C}$ to a category $\mathcal{D}$ is specified by the following:
            \begin{itemize}
                \item 
                    a function 
                    \[\begin{aligned}
                        \textbf{obj}(\mathcal{C}) &\to \textbf{obj}(\mathcal{D}) \\
                        X &\mapsto F(X)
                    \end{aligned}\]

@@

        \subsection{Natural transformation}
        Before introducing the definition of natural transformation, we need to get familiar with the notation of commutative diagrams.

        \subsection{Presheaf}

        \subsection{Functor category}

        \subsection{Cartesian closed category}

        \subsection{Yoneda lemma}

        \subsection{Functor categories}

    \section{Agda}

@@

    The interactive development of Agda with holes and type checking is very useful for debugging.

@@

    \begin{enumerate}
        \item 
            Tools used:
            git version control

            have tried emacs
            vscode interactive 
    \end{enumerate}

    \printbibliography
\end{document}

>>>>

\graphicspath{{./fig/}}

@@

    \section{Category theory background}
        Category 
        \subsection{Category}
        \begin{definition*}[Category]
            A \emph{category} $\mathcal{C}$ is specified by the following:
            \begin{itemize}
                \item 
                    a collection of objects $\textbf{obj}(\mathcal{C})$, whose elements are called $\mathcal{C}$-objects;
                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a collection of morphisms $\mathcal{C}{(X,Y)}$, whose elements are called $\mathcal{C}$-morphisms from $X$ to $Y$;
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, an element $\textbf{id}_X \in \mathcal{C}{(X,X)}$ called the identity morphism on $X$;
                \item 
                    for each $X, Y, Z \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} \times \mathcal{C}{(Y,Z)} &\to \mathcal{C}{(X,Z)} \\
                        (f,g) &\mapsto g \circ f
                    \end{aligned}\]
                    called the composition of morphisms;
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    \textbf{(Unit Laws)}
                    for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have:
                    \begin{equation} \label{law: unit}
                        \textbf{id}_Y \circ f = f = f \circ \textbf{id}_X
                    \end{equation}
                \item
                    \textbf{(Associativity Law)}
                    for all $X, Y, Z, W \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, $h \in \mathcal{C}{(Z,W)}$, we have:
                    \begin{equation} \label{law: associativity}
                        h \circ (g \circ f) = (h \circ g) \circ f
                    \end{equation}
            \end{itemize}

            [Considering adding different notations for morphisms, composition, identity morphisms, etc.]

            [Considering adding a diagram for the definition of category]

            [Considering adding examples of categories]

@@

        Before other definitions, we need to get familiar with the notation of commutative diagrams.
        \subsubsection{Commutative diagrams}
        A \emph{diagram} in a category $\mathcal{C}$ is a directed graph whose vertices are $\mathcal{C}$-objects and whose edges are $\mathcal{C}$-morphisms.

        A diagram is \emph{commutative} (or \emph{commutes}) if any two finite paths in the graph between any two vertices $X$ and $Y$ in the diagram determine the equal morphism $f \in \mathcal{C}{(X,Y)}$ under the composition of morphisms.

        \begin{figure}[htb!]
            \centering
            \begin{minipage}{0.3\textwidth}
                \centering
                \includegraphics[width=0.8\textwidth]{2_2_1_unit_laws.pdf}
                \caption{Commutative diagram for Unit Laws}
                \label{fig: unit_laws}
            \end{minipage} \hspace{0.2\textwidth}
            \begin{minipage}{0.3\textwidth}
                \centering
                \includegraphics[width=0.8\textwidth]{2_2_1_associativity_law.pdf}
                \caption{Commutative diagram for Associativity Law}
                \label{fig: associativity_law}
            \end{minipage}
        \end{figure}

        As examples of commutative diagrams, Figure \ref{fig: associativity_law} and Figure \ref{fig: unit_laws} are commutative diagrams for the unit laws and associativity law respectively.

        \subsection{Cartesian closed category}
        \subsection{Functor}
        \begin{definition*}[Functor]
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, a \emph{functor} $F: \mathcal{C} \to \mathcal{D}$ is specified by:
            \begin{itemize}
                \item 
                    a function 
                    \[\begin{aligned}
                        \textbf{obj}(\mathcal{C}) &\to \textbf{obj}(\mathcal{D}) \\
                        X &\mapsto F(X)
                    \end{aligned}\]

@@

        \subsection{Natural transformation}
        \begin{definition*}[Natural transformation]

            Given two categories $\mathcal{C}$ and $\mathcal{D}$, and two functors $F,G: \mathcal{C} \to \mathcal{D}$, a \emph{natural transformation} $\theta : F \to G$ is a family of morphisms $\theta_X \in \mathcal{D}{(F(X),G(X))}$ for each $X \in \textbf{obj}(\mathcal{C})$ such that for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, the following diagram 
            \begin{equation}
                \vcenter{\hbox{\includegraphics[width=0.3\textwidth]{2_2_3_natural_transformation.pdf}}}
                \label{cd: natural_transformation}
            \end{equation}

            commutes in $\mathcal{D}$, i.e. the following equation holds:
            \begin{equation} \label{law: natural_transformation}
                G(f) \circ \theta_X = \theta_Y \circ F(f)
            \end{equation}

        \end{definition*}
        [Considering adding examples of natural transformations]

        \subsection{Functor category}
        \begin{definition*}[Functor category]
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, the \emph{functor category} $\mathcal{D}^{\mathcal{C}}$ is the category satisfying the following:
            \begin{itemize}
                \item 
                    the objects of $\mathcal{D}^{\mathcal{C}}$ are all functors $\mathcal{C} \to \mathcal{D}$;
                \item 
                    given two functors $F,G: \mathcal{C} \to \mathcal{D}$, the morphisms from $F$ to $G$ in $\mathcal{D}^{\mathcal{C}}$ are all natural transformations $\theta : F \to G$;
                \item
                    composition and identity morphisms in $\mathcal{D}^{\mathcal{C}}$ are defined as follows:
                    \begin{itemize}
                        \item 
                            the identity morphism $\textbf{id}_F$ on $F$ is defined as $\theta_X = \textbf{id}_{F(X)}$ for all $X \in \textbf{obj}(\mathcal{C})$;
                        \item
                            the composition of two natural transformations $\theta : F \to G$ and $\phi : G \to H$ is defined as $(\phi \circ \theta)_X = \phi_X \circ \theta_X$ for all $X \in \textbf{obj}(\mathcal{C})$.
                    \end{itemize}
            \end{itemize}
        \end{definition*}

        \subsection{Yoneda lemma}
        \begin{definition*}[Yoneda functor]
            Given a category $\mathcal{C}$, the \emph{Yoneda functor} $y: \mathcal{C} \to \mathcal{C}^{\mathcal{C}^{op}}$ is defined as follows:
            \begin{itemize}
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, $y(X)$ is the functor $y(X): \mathcal{C}^{op} \to \textbf{Set}$ defined as:
                    \[ y(X)(Y) = \mathcal{C}{(Y,X)} \]
                    for all $Y \in \textbf{obj}(\mathcal{C})$;
                \item
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $y(f)$ is the natural transformation $y(f): y(X) \to y(Y)$ defined as:
                    \[ y(f)_Y = f \circ - : y(X)(Y) \to y(Y)(Y) \]
                    for all $Y \in \textbf{obj}(\mathcal{C})$.
                    In other words, $y(f)_Y$ is the function that takes an element $g \in y(X)(Y) = \mathcal{C}{(Y,X)}$ and returns the element $f \circ g \in y(Y)(Y) = \mathcal{C}{(Y,Y)}$.
                    The function $y(f)_Y$ is a morphism in $\mathcal{C}^{\mathcal{C}^{op}}$ from $y(X)(Y)$ to $y(Y)(Y)$.
                    In other words, $y(f)_Y$ is a morphism in $\mathcal{C}^{\mathcal{C}^{op}}$ from the functor $y(X)$ to the functor $y(Y)$.
                    The morphism $y(f)_Y$ is defined for all $Y \in \textbf{obj}(\mathcal{C})$.
                    In other words, $y(f)$ is a natural transformation from the functor $y(X)$ to the functor $y(Y)$.
            \end{itemize}
        \end{definition*}

        \begin{definition*}[Yoneda lemma]
            Given a category $\mathcal{C}$, the \emph{Yoneda lemma} states that for any functor $F: \mathcal{C} \to \textbf{Set}$, there is a natural isomorphism:
            \[ \begin{aligned}
                \textbf{Nat}(y(X),F) &\cong F(X) \\
                \theta &\mapsto \theta_X
            \end{aligned} \]
            for all $X \in \textbf{obj}(\mathcal{C})$, where $\textbf{Nat}(y(X),F)$ is the set of natural transformations from the functor $y(X)$ to the functor $F$.
        \end{definition*}

        \subsection{Presheaf}
        \begin{definition*}[Presheaf]
            Given a category $\mathcal{C}$, a \emph{presheaf} on $\mathcal{C}$ is a functor $F: \mathcal{C}^{op} \to \textbf{Set}$.
            A presheaf is a contravariant functor, which means that it reverses the direction of morphisms.
            In other words, a presheaf is a functor that takes objects in $\mathcal{C}$ and assigns them sets, and takes morphisms in $\mathcal{C}$ and assigns them functions between the corresponding sets.
            The presheaf $F$ is defined as follows:
            \begin{itemize}
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, $F(X)$ is a set;
                \item
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $F(f)$ is a function $F(Y) \to F(X)$.
                    In other words, $F(f)$ is a function that takes an element $g \in F(Y)$ and returns the element $f \circ g \in F(X)$.
                    The function $F(f)$ is a morphism in $\mathcal{C}^{op}$ from $F(Y)$ to $F(X)$.
                    In other words, $F(f)$ is a morphism in $\mathcal{C}^{op}$ from the functor $F(Y)$ to the functor $F(X)$.
                    The morphism $F(f)$ is defined for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$.
                    In other words, $F(f)$ is a natural transformation from the functor $F(Y)$ to the functor $F(X)$.
            \end{itemize}
        \end{definition*}

        \subsection{Cartesian closed structure in presheaf categories}

    \section{Agda}
        \subsection{Standard library}

        \subsection{Dependent Types}

        \subsection{Curry-Howard Correspondence}

        \subsection{Equality}

        \subsection{Proof Example}

        \subsection{Interactive programming with holes}

@@

@@

\chapter{Implementation}

\chapter{Evaluation}

\chapter{Conclusion}

    \printbibliography

\end{document}
